\bf Landslide: \\ Systematic Dynamic Race Detection in Kernel Space

### 2012-05-16 design.tex
'fix' figure in terminology section
more minor
fix some footnotes

%\subsection{Stack Tracing}
% FIXME

@@

\begin{itemize}
	\item {\bf Always-accessed memory locations.}
		In kernel-space, every thread switch goes through common scheduler and context-switcher routines. These routines inevitably access scheduler data structures, such as the runqueue.\footnote{
		Another example is the timer tick counter: this global variable is not necessarily accessed every transition, because voluntary reschedules do not involve timer interrupts, but every involuntary preemption will cause a write to this counter.}
		If we include such memory accesses in the independence relation, it will result that all transitions conflict, and DPOR won't be able to achieve any reduction. To allow for state space reduction, we sacrifice our ability to find races involving these particular accesses by ignoring them. Currently we require the user to identify these locations as part of the instrumentation process, as described in Section~\ref{sec:using-config-landslide}.
	\item {\bf Freed memory poisoning.}
		A use-after-free bug happens whenever one thread accesses an address within an allocated block that another thread previously freed. Even if no other code makes a conflicting access to the same address after the \texttt{free}, accessing dynamically-allocated memory after it has been freed is illegal no matter what.

@@

\newcommand\hilight[2]{\color{#1}#2\color{white}}
\begin{figure}[h]
	%\includegraphics[width=0.95\textwidth]{found_a_bug.png}
	\centering
	\colorbox{black}{\color{white}
	{\small
	\begin{tabular}{l}
	\texttt{\hilight{green}{[SCHEDULE]}~~~~~~thread 4 vanished} \\
	\texttt{\hilight{green}{[SCHEDULE]}~~~~~~switched threads 4 -> 3} \\
	\texttt{\hilight{brown}{[MEMORY]}~~~~~~~~\hilight{red}{USE AFTER FREE - read from 0x0015a8f0 at eip 0x00104209}} \\
	\texttt{\hilight{brown}{[MEMORY]}~~~~~~~~Heap contents:~\{...\}} \\
	\texttt{\hilight{brown}{[MEMORY]}~~~~~~~~[0x15a8f0 | 4136] was allocated by TID3 at (...)} \\
	\texttt{\hilight{brown}{[MEMORY]}~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~and freed by TID4 at (...)} \\
	\texttt{\hilight{red}{[BUG!]}~~~~~~~~~~\hilight{red}{****~~~~~A bug was found!~~~~****}} \\
	\texttt{\hilight{red}{[BUG!]}~~~~~~~~~~\hilight{red}{**** Decision trace follows.~****}} \\
	\texttt{\hilight{red}{[BUG!]}~~~~~~~~~~\hilight{yellow}{1:~~1347079 instructions, old 3 new 4, current 4}} \\
	\texttt{\hilight{red}{[BUG!]}~~~~~~~~~~~~~~TID3 at 0x00105a10 in \hilight{cyan}{context\_switch},} \\
	\texttt{\hilight{red}{[BUG!]}~~~~~~~~~~~~~~~~~~~~~~0x001041f4 in \hilight{cyan}{thread\_fork},} \\
	\texttt{\hilight{red}{[BUG!]}~~~~~~~~~~~~~~~~~~~~~~0x0010362b in \hilight{cyan}{thread\_fork\_wrapper}} \\
	\texttt{\hilight{red}{[BUG!]}~~~~~~~~~~\hilight{yellow}{2:~~1350725 instructions, old 4 new 3, current 3}} \\
	\texttt{\hilight{red}{[BUG!]}~~~~~~~~~~~~~~TID4 at 0x00105a10 in \hilight{cyan}{context\_switch},} \\
	\texttt{\hilight{red}{[BUG!]}~~~~~~~~~~~~~~~~~~~~~~0x00104681 in \hilight{cyan}{yield},} \\
	\texttt{\hilight{red}{[BUG!]}~~~~~~~~~~~~~~~~~~~~~~0x00104570 in \hilight{cyan}{vanish},} \\
	\texttt{\hilight{red}{[BUG!]}~~~~~~~~~~~~~~~~~~~~~~0x00103708 in \hilight{cyan}{vanish\_wrapper}} \\
	\texttt{\hilight{red}{[BUG!]}~~~~~~~~~~Stack:~TID3 at 0x00104209 in \hilight{cyan}{thread\_fork},} \\
	\texttt{\hilight{red}{[BUG!]}~~~~~~~~~~~~~~~~~~~~~~~~~0x0010362b in \hilight{cyan}{thread\_fork\_wrapper}} \\
	\texttt{\hilight{red}{[BUG!]}~~~~~~~~~~Total decision points 24, total backtracks 5} \\
	\texttt{\hilight{red}{[BUG!]}~~~~~~~~~~Average instrs/decision 16155, average branch depth 5} \\
	\end{tabular}
	}
	}
	\caption{A sample decision trace (with extraneous text trimmed) that Landslide generated using the \texttt{double\_thread\_fork} test case (Section~\ref{sec:using-tests}).}
	\label{fig:found_a_bug}
\end{figure}

>>>>

@@

\begin{itemize}
	\item {\bf Always-accessed memory locations.}
		In kernel-space, every thread switch goes through common scheduler and context-switcher routines. These routines inevitably access scheduler data structures, such as the runqueue. The timer tick counter is another example: this global variable is not necessarily accessed every transition, because voluntary reschedules do not involve timer interrupts, but every involuntary preemption will cause a write to this counter.

		If we include such memory accesses in the independence relation, it will result that all transitions conflict, and DPOR won't be able to achieve any reduction. To allow for state space reduction, we sacrifice our ability to find races involving these particular accesses by ignoring them. Currently we require the user to identify these locations as part of the instrumentation process, as described in Section~\ref{sec:using-config-landslide}.
	\item {\bf Freed memory poisoning.}
		A use-after-free bug happens whenever one thread accesses an address within an allocated block that another thread previously freed. Even if no other code makes a conflicting access to the same address after the \texttt{free}, accessing dynamically-allocated memory after it has been freed is illegal no matter what.

@@

\begin{figure}[h]
	%\includegraphics[width=0.95\textwidth]{found_a_bug.png}
	\centering
	\colorbox{black}{\color{white}
	{\small
	\begin{tabular}{l}
	\texttt{\hilight{green}{[SCHEDULE]}~~~~~~thread 4 vanished} \\
	\texttt{\hilight{green}{[SCHEDULE]}~~~~~~switched threads 4 -> 3} \\
	\texttt{\hilight{brown}{[MEMORY]}~~~~~~~~\hilight{red}{USE AFTER FREE - read from 0x0015a8f0 at eip 0x00104209}} \\
	\texttt{\hilight{brown}{[MEMORY]}~~~~~~~~Heap contents:~\{...\}} \\
	\texttt{\hilight{brown}{[MEMORY]}~~~~~~~~[0x15a8f0 | 4136] was allocated by TID3 at (...)} \\
	\texttt{\hilight{brown}{[MEMORY]}~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~and freed by TID4 at (...)} \\
	\texttt{\hilight{red}{[BUG!]}~~~~~~~~~~\hilight{red}{****~~~~~A bug was found!~~~~****}} \\
	\texttt{\hilight{red}{[BUG!]}~~~~~~~~~~\hilight{red}{**** Decision trace follows.~****}} \\
	\texttt{\hilight{red}{[BUG!]}~~~~~~~~~~\hilight{yellow}{1:~~1347079 instructions, old 3 new 4, current 4}} \\
	\texttt{\hilight{red}{[BUG!]}~~~~~~~~~~~~~~TID3 at 0x00105a10 in \hilight{cyan}{context\_switch},} \\
	\texttt{\hilight{red}{[BUG!]}~~~~~~~~~~~~~~~~~~~~~~0x001041f4 in \hilight{cyan}{thread\_fork},} \\
	\texttt{\hilight{red}{[BUG!]}~~~~~~~~~~~~~~~~~~~~~~0x0010362b in \hilight{cyan}{thread\_fork\_wrapper}} \\
	\texttt{\hilight{red}{[BUG!]}~~~~~~~~~~\hilight{yellow}{2:~~1350725 instructions, old 4 new 3, current 3}} \\
	\texttt{\hilight{red}{[BUG!]}~~~~~~~~~~~~~~TID4 at 0x00105a10 in \hilight{cyan}{context\_switch},} \\
	\texttt{\hilight{red}{[BUG!]}~~~~~~~~~~~~~~~~~~~~~~0x00104681 in \hilight{cyan}{yield},} \\
	\texttt{\hilight{red}{[BUG!]}~~~~~~~~~~~~~~~~~~~~~~0x00104570 in \hilight{cyan}{vanish},} \\
	\texttt{\hilight{red}{[BUG!]}~~~~~~~~~~~~~~~~~~~~~~0x00103708 in \hilight{cyan}{vanish\_wrapper}} \\
	\texttt{\hilight{red}{[BUG!]}~~~~~~~~~~Stack:~TID3 at 0x00104209 in \hilight{cyan}{thread\_fork},} \\
	\texttt{\hilight{red}{[BUG!]}~~~~~~~~~~~~~~~~~~~~~~~~~0x0010362b in \hilight{cyan}{thread\_fork\_wrapper}} \\
	\texttt{\hilight{red}{[BUG!]}~~~~~~~~~~Total decision points 24, total backtracks 5} \\
	\texttt{\hilight{red}{[BUG!]}~~~~~~~~~~Average instrs/decision 16155, average branch depth 5} \\
	\end{tabular}
	}
	}
	\caption{A sample decision trace (with extraneous text trimmed) that Landslide generated using the \texttt{double\_thread\_fork} test case (Section~\ref{sec:using-tests}).}
	\label{fig:found_a_bug}
\end{figure}

### 2012-05-16 evaluation.tex
fix some footnotes

Finding this bug was somewhat of a surprise. When two threads in the same process wait for a child to exit, they both block on the same condition variable, and one of them has a pointer to the other in its \texttt{cond\_queue\_next} pointer. However, when awakened, the condition variable code did not clear the value of this pointer. Fortunately, an assert statement in the thread destruction code caught that the pointer was non-\texttt{NULL} when it should have been, and the kernel panicked.\footnote{
Even with this assert, extremely mysterious behaviour could have arisen in other use cases, such as during stress testing: if the awakened thread had gone to sleep on another condition variable, its \texttt{cond\_queue\_next} pointer could point to a thread asleep on a different condition variable, in which case a broadcast on the new condition variable cause a spurious wakeup of the second thread. Worse yet, if the second thread had also been awoken, and thence exited, a broadcast would result in a use-after-free.}

@@

This bug is exposed with the \texttt{double\_thread\_fork} test case. The complexity is relatively simple\footnote{
Many userspace thread libraries forcibly serialise invocations of \texttt{thread\_fork} and \texttt{vanish} with their own synchronisation barriers, so the kernel bug may never be noticed when using such libraries (this caused the author some trouble).},
and the fix is only two lines long; the dereference should simply occur before making the child thread runnable.

>>>>

Finding this bug was somewhat of a surprise. When two threads in the same process wait for a child to exit, they both block on the same condition variable, and one of them has a pointer to the other in its \texttt{cond\_queue\_next} pointer. However, when awakened, the condition variable code did not clear the value of this pointer. Fortunately, an assert statement in the thread destruction code caught that the pointer was non-\texttt{NULL} when it should have been, and the kernel panicked.

Even with this assert, extremely mysterious behaviour could have arisen in other use cases, such as during stress testing: if the awakened thread had gone to sleep on another condition variable, its \texttt{cond\_queue\_next} pointer could point to a thread asleep on a different condition variable, in which case a broadcast on the new condition variable would cause a spurious wakeup of the second thread. Worse yet, if the second thread had also been awoken, and thence exited, a broadcast would result in scheduler data structure corruption.

@@

This bug is exposed with the \texttt{double\_thread\_fork} test case. The complexity is relatively simple,\footnote{
Many userspace thread libraries forcibly serialise invocations of \texttt{thread\_fork} and \texttt{vanish} with their own synchronisation barriers, so the kernel bug may never be noticed when using such libraries (this caused the author some trouble).}
and the fix is only two lines long; the dereference should simply occur before making the child thread runnable.

### 2012-05-16 terminology.tex
fix some footnotes

\resume{enumerate}
	\item {\bf Thread}:
		One participating agent in a concurrent program, which executes code sequentially, with the potential to be interleaved with the execution of other threads in the system.
		Each thread has a unique numeric identifier (TID).
	\item {\bf Involuntary preemption}:
		A context switch from one thread to another caused by a nondeterministic event, such as a timer tick or device interrupt. The systematic testing framework must control all sources of nondeterminism. In this work, we focus only on timer-driven thread switches.
	\item {\bf Voluntary reschedule}:
		A context switch from one thread to another which the guest kernel performed automatically, not triggered by nondeterministic events and/or Landslide but rather as part of its normal execution.\footnote{Voluntary reschedules may or may not be necessary for the kernel to behave correctly at all (for example: switching away from an exiting thread or from a thread blocked on \texttt{wait} is necessary, but switching to a newly-forked thread is not). Involuntary preemptions are never necessary for correct kernel behaviour.}
		See Section~\ref{sec:components-arbiter}.
\suspend{enumerate}

>>>>

\resume{enumerate}
	\item {\bf Thread}:
		One participating agent in a concurrent program, which executes code sequentially, with the potential to be interleaved with the execution of other threads in the system.
		Each thread has a unique numeric identifier (TID).
	\item {\bf Involuntary preemption}:
		A context switch from one thread to another caused by a nondeterministic event, such as a timer tick or device interrupt. The systematic testing framework must control all sources of nondeterminism. In this work, we focus only on timer-driven thread switches.
	\item {\bf Voluntary reschedule}:
		A context switch from one thread to another which the guest kernel performed automatically, not triggered by nondeterministic events and/or Landslide but rather as part of its normal execution.

		Voluntary reschedules may or may not be necessary for the kernel to behave correctly at all (for example: switching away from an exiting thread or from a thread blocked on \texttt{wait} is necessary, but switching to a newly-forked thread is not). Involuntary preemptions, however, are never necessary for correct kernel behaviour.
		See Section~\ref{sec:components-arbiter}.
\suspend{enumerate}

### 2012-05-16 using.tex
found_a_bug in text
fix some footnotes

\newcommand\hilight[2]{\color{#1}#2\color{white}}
\begin{figure}[h]
	\centering
	\colorbox{black}{\color{white}
	{\small
	\begin{tabular}{l}
	\texttt{\hilight{green}{int}~thread\_fork()} \\
	\texttt{\{} \\
	\texttt{~~~~\hilight{green}{thread\_t}~*child = construct\_new\_thread();} \\
	\texttt{~~~~\hilight{cyan}{tell\_landslide\_forking}();} \\
	\texttt{~~~~add\_to\_runqueue(child);} \\
	\texttt{~~~~\hilight{cyan}{tell\_landslide\_decide}(); \hilight{purple}{/* Interrupt me here! */}} \\
	\texttt{~~~~\hilight{brown}{return}~child->tid;} \\
	\texttt{\}} \\
	\end{tabular}
	}
	}
	\caption{Example annotated code. In theory, the use of \texttt{tell\_landslide\_decide()} in the place where a preemption needs to occur is not necessary for Landslide to know to preempt then; it is shown for the sake of demonstration.}
	\label{fig:tell-landslide}
\end{figure}

@@

\begin{itemize}
	\small
	\item \texttt{sched\_func}: When functions that make up the kernel's scheduler are identified, Landslide knows to ignore shared memory accesses from them. Accesses to scheduler data structures happen during {\em every} transition, so ignoring these is necessary for achieving any reduction at all.
	\item \texttt{ignore\_sym}: The user may wish to ignore memory accesses from other global data structures as well. For example, ignoring a mutex which is locked very frequently, but irrelevant to the actual test case, will result in an independence relation that \revision{enables much better state space reduction without losing the ability to find bugs in the parts of the kernel the user is interested in testing}.
	\item \texttt{within\_function}: If the user configures decision points to happen in common code paths, such as \texttt{mutex\_lock}, it will likely generate more branching than needed.\footnote{
For example, if testing \texttt{vanish}, there's no need to branch on a \texttt{mutex\_lock} in \texttt{exec}.}
This configuration option allows the user to whitelist functions so that Landslide will only decide if the current thread is ``within'' one of those functions.
	\item \texttt{without\_function}: Similar to above, but a blacklist. These two can be used together; later invocations take precedence over earlier ones.
\end{itemize}

>>>>

\begin{figure}[h]
	\centering
	\colorbox{black}{\color{white}
	{\small
	\begin{tabular}{l}
	\texttt{\hilight{green}{int}~thread\_fork()} \\
	\texttt{\{} \\
	\texttt{~~~~\hilight{green}{thread\_t}~*child = construct\_new\_thread();} \\
	\texttt{~~~~\hilight{cyan}{tell\_landslide\_forking}();} \\
	\texttt{~~~~add\_to\_runqueue(child);} \\
	\texttt{~~~~\hilight{cyan}{tell\_landslide\_decide}(); \hilight{magenta}{/* Interrupt me here! */}} \\
	\texttt{~~~~\hilight{brown}{return}~child->tid;} \\
	\texttt{\}} \\
	\end{tabular}
	}
	}
	\caption{Example annotated code. In theory, the use of \texttt{tell\_landslide\_decide()} in the place where a preemption needs to occur is not necessary for Landslide to know to preempt then; it is shown for the sake of demonstration.}
	\label{fig:tell-landslide}
\end{figure}

@@

\begin{itemize}
	\small
	\item \texttt{sched\_func}: When functions that make up the kernel's scheduler are identified, Landslide knows to ignore shared memory accesses from them. Accesses to scheduler data structures happen during {\em every} transition, so ignoring these is necessary for achieving any reduction at all.
	\item \texttt{ignore\_sym}: The user may wish to ignore memory accesses from other global data structures as well. For example, ignoring a mutex which is locked very frequently, but irrelevant to the actual test case, will result in an independence relation that \revision{enables much better state space reduction without losing the ability to find bugs in the parts of the kernel the user is interested in testing}.
	\item \texttt{within\_function}: If the user configures decision points to happen in common code paths, such as \texttt{mutex\_lock}, it will likely generate more branching than needed. For example, if testing \texttt{vanish}, there's no need to branch on a \texttt{mutex\_lock} in \texttt{exec}.

This configuration option allows the user to whitelist functions so that Landslide will only decide if the current thread is ``within'' one of those functions.
	\item \texttt{without\_function}: Similar to above, but a blacklist. These two can be used together; later invocations take precedence over earlier ones.
\end{itemize}
